---
title: "Explorative Analysis"
format: 
  html:
    toc: true
    code-fold: show
    df-print: paged
    warnings: false
---

## Load Libraries

```{r}
pacman::p_load(
  tidyverse,  # A set of many useful libraries
  readxl,     # To import the dataset from Excel
  here,       # To avoid problems with file directories
  janitor,    # To clean data in a fast way
  gt,         # Output tables
  gtsummary,  # Output tables for models and survival data
  patchwork # merge more plots
)
```


Set ggplot theme

```{r}
theme_set(theme_minimal())
```


## Import Data

```{r}
data <- read_csv(
  here("data", "FINAL", "dataframe_cleaned.csv")
  )

data |> 
  slice_sample(n = 1, by = c(donation_type, gender)) |> 
  rename_all(\(x) str_replace_all(x, "_", " ")) |> 
  gt() |> 
  tab_header("Sample of the dataset cleaned",
             "The sample was stratified by the variables donation_type and gender")
```

## Some Checks

Check the unique donors number

```{r}
data |> 
  distinct(unique_number) |> 
  nrow()
```
Check if there are any duplicate rows

```{r}
data |> 
  janitor::get_dupes() 
```
### Expand the rows

Add the observation for the years which the donators haven't donated.

```{r}
data |>
  complete(unique_number, year = full_seq(year, 1), fill = list(number_of_donations = 0)) |>
  mutate(
    across(c(gender, birth_year, birth_cohort, first_donation_year, first_donation_cohort, donation_type, donor_class),
           \(x) coalesce(x, first(na.omit(x)))),
    age = year - birth_year,
    .by = unique_number
  ) -> data_with_zeros

data_with_zeros |> 
  slice_min(n = 1, order_by = unique_number) |> 
  select(unique_number, year, age, number_of_donations)
```



### Clean the dfs

```{r}
data |> 
  distinct() |> 
  mutate(
    class_year = cut(birth_year, 
                     breaks = seq(1900, 2010, by = 10), 
                     dig.lab = 4,
                     include.lowest = TRUE
                     ),
    class_age = cut(age, 
                     breaks = c(seq(0, 70, by = 10), max(age)), 
                     dig.lab = 3,
                     include.lowest = TRUE
                     ),
    .before = birth_year
  ) -> data

data_with_zeros |> 
  distinct() |> 
  mutate(
    class_year = cut(birth_year, 
                     breaks = seq(1900, 2010, by = 10), 
                     dig.lab = 4,
                     include.lowest = TRUE
                     ),
    class_age = cut(age, 
                     breaks = c(seq(0, 70, by = 10), max(age)), 
                     dig.lab = 3,
                     include.lowest = TRUE
                     ),
    .before = birth_year
  ) -> data_with_zeros
```


### Summarised df

```{r}
data |> 
  summarise(
    number_of_donations = sum(number_of_donations),
    .by = -c(number_of_donations, unique_number)
  ) -> summarised_data
```



## Some analysis

Age of donors for the last year

```{r}
data |> 
  filter(
    year == max(year) | year == min(year)
  ) |> 
  ggplot(aes(x = gender, y = age, fill = gender)) +
  geom_violin() +
  facet_wrap(~year) +
  labs(
    title = "Donor age distribution in 2009 and 2023 by gender"
    )
```


```{r}
data |> 
  select(class_year, donation_type) |> 
  tbl_summary(
    by = donation_type
  ) |> 
  add_overall()
```

```{r}
data |> 
  reframe(
    class_year, donor_class,
    .by = unique_number
  ) |> 
  select(class_year, donor_class) |> 
  tbl_summary(
    by = donor_class
  ) |> 
  add_overall()
```
Maybe the donations are reffered to the year. But is it possible to donate 17 times per year?

```{r}
data |> 
  tabyl(number_of_donations, gender) |>
  adorn_percentages(denominator = "col") |>  
  adorn_pct_formatting()
```



```{r}
data |> 
  tabyl(number_of_donations, donor_class) |>
  adorn_percentages(denominator = "col") |>  
  adorn_pct_formatting()
```

## Time series

```{r}
data |>
  filter(
    donation_type == "SANGUE",
    birth_year > 1930
    ) |> 
  summarise(
    donations = sum(number_of_donations),
    .by = c(class_year, year)
  ) |> 
  ggplot(aes(year, donations, color = class_year, group = class_year)) +
  geom_line() +
  scale_color_brewer() +
  labs(
    title = "Number of donations by year and class of birth years"
  )
```
```{r}
data |>
  filter(
    donation_type == "SANGUE",
    birth_year > 1930
    ) |> 
  summarise(
    donations = sum(number_of_donations),
    .by = c(class_age, year)
  ) |> 
  ggplot(aes(year, donations, color = class_age, group = class_age)) +
  geom_line() +
  scale_color_brewer(direction = -1) +
  labs(
    title = "Number of donations by year and class of birth years"
  )
```

```{r}
data |>
  summarise(
    donations = sum(number_of_donations),
    .by = c(donation_type, year)
  ) |> 
  ggplot(aes(year, donations, color = donation_type, group = donation_type)) +
  geom_line() +
  geom_smooth(linetype = "dashed",
              alpha = .15,
              size = .7
              ) +
  scale_color_brewer(type = "qual", palette = 2) +
  labs(
    title = "Number of donations by year and donation type"
  )
```

```{r}
data |>
  summarise(
    donations = sum(number_of_donations),
    .by = c(donor_class, year)
  ) |> 
  ggplot(aes(year, donations, color = donor_class, group = donor_class)) +
  geom_line() +
  geom_smooth(linetype = "dashed",
              alpha = .15,
              size = .7
              ) +
  scale_color_brewer(type = "qual", palette = 2) +
  labs(
    title = "Number of donations by year and donor class"
  )
```
## Integrate Data


Integrate data with Trieste residents to get more insights

```{r}
residenti <- read_csv(here("data", "residenti_Trieste_e_Gorizia.csv"))

residenti |> 
  filter(
    Sesso != "totale"
  ) |> 
  rename(
    age = Età,
    gender = Sesso,
    year = TIME,
    population = Value,
  ) |> 
  summarise(
    across(population, sum),
    .by = c(age, gender, year)  
  ) |> 
  mutate(
    age = str_extract(age, "[:digit:]*") |> as.numeric(),
    gender = case_when(
      gender == "maschi" ~ "M",
      gender == "femmine" ~ "F",
      T ~ NA
    )
  ) -> residenti
```


### Estimating Past Residents

For the years prior to 2019, one could use the 2019 data from ISTAT, adjust the ages, and account for population changes with mortality tables, assuming a constant flow of immigrants and emigrants.

```{r}
lifecontingencies::demoIta |> 
  transmute(
    age = X,
    maschi = SIM02,
    femmine = SIF02,
    across(c(maschi, femmine), \(x) x / lag(x, default = 1e5), .names = "{col}_px")
  ) |> 
  pivot_longer(ends_with("px"), names_to = "gender", values_to = "px") |>
  # select(-maschi, - femmine) |> 
  mutate(
    gender = case_when(
      gender == "maschi_px" ~ "M",
      gender == "femmine_px" ~ "F",
      T ~ NA
    )
  ) -> life_table

residenti |> 
  add_row(year = 2009:2018, .before = 1) |> 
  complete(age, gender, year) |> 
  filter(!if_any(c(age, gender), is.na)) |> 
  left_join(life_table, by = c("age", "gender")) |> 
  arrange(-year, age) |>
  filter(gender == "M") |> 
  pivot_wider(names_from = year, values_from = population, names_prefix = "y_") |> 
  mutate(
    y_2018 = lead(y_2019, default = 0) * maschi / lead(maschi, default = 0),
    y_2017 = lead(y_2019, n = 2, default = 0) * maschi / lead(maschi, default = 0),
    y_2016 = lead(y_2019, n = 3, default = 0) * maschi / lead(maschi, default = 0),
  )
```

Let $l_x$ be the population from the mortality table at age $x$, and let $n_x^y$ be the resident population in Trieste in year $y$ at age $x$. To estimate the population in year $y_j$ at age $x$, given the population in year $y_i$, the following formula is used:

$n^{y_i}_x = n^{y_j}_{x - (y_i - y_j)} \cdot \frac{l_x}{l_{x - (y_i - y_j)}}$

```{r}
filled_residenti <-
  residenti |> 
  add_row(year = 2009:2018, .before = 1) |> 
  complete(age, gender, year) |> 
  filter(!if_any(c(age, gender), is.na)) |> 
  left_join(life_table, by = c("age", "gender")) |> 
  arrange(-year, age) |>
  pivot_wider(names_from = gender, values_from = c(px, population)) |> 
  pivot_wider(names_from = year, values_from = c(population_M, population_F))

years <- 2018:2009
gender <- "M"

filled_residenti <- reduce(years, function(df, year) {
  col <- paste0("population_", c("F", "M"), "_", year)
  col_19 <- paste0("population_", c("F", "M"), "_2019")
  df |> 
    mutate(
      !!sym(col[1]) := lead(!!sym(col_19[1]), 2019 - year, 0) * femmine / lead(femmine, 2019 - year, 0),
      !!sym(col[2]) := lead(!!sym(col_19[2]), 2019 - year, 0) * maschi / lead(maschi, 2019 - year, 0)
      )
}, .init = filled_residenti) |> 
  mutate(across(starts_with("population"), round))

filled_residenti
```

```{r}
filled_residenti |> 
  select(age, starts_with("population")) |> 
  pivot_longer(starts_with("population"), values_to = "population") |> 
  separate(name, sep = "_", into = c("pop", "gender", "year")) |> 
  select(-pop) |> 
  mutate(across(year, as.numeric)) -> residenti
```


### Plot Integrated Data

```{r}
data |> 
  left_join(residenti, by = c("gender", "year", "age")) |>
  filter(year %% 4 == 3) |>
  summarise(
    ratio_donors = n() / first(population),
    .by = c(gender, year, age)
  ) |> 
  filter(age < 75) |> 
  ggplot(aes(x = age, y = ratio_donors, fill = gender, alpha = year)) +
  geom_col() +
  facet_grid(rows = vars(year), cols = vars(gender)) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_alpha_continuous(range = c(.4, 1)) +
  labs(
    title = "Donors population ratio by gender and year"
  ) +
  theme(legend.position = "none")
```


```{r}
data |> 
  left_join(residenti, by = c("gender", "year", "age")) |>
  reframe(
    n = n(),
    .by = c(gender, year, age, class_age, population)
  ) |> 
  summarise(
    ratio_donors = sum(n) / sum(population),
    .by = c(year, gender, class_age)
  ) |> 
  ggplot(aes(x = year, y = ratio_donors, color = class_age)) +
  geom_line() +
  facet_grid(cols = vars(gender)) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Donors population ratio by gender and year"
  )
```

# Statistical models

## Analyze the dependent variable


```{r}
#| fig-width: 6
#| fig-height: 8

map(
  c("donor_class", "donation_type", "gender"),
  \(x) {
    data |> 
      summarise(
        n = n(),
        .by = c(!!sym(x), number_of_donations)
      ) |>
      ggplot(aes(x = number_of_donations, y = n, fill = !!sym(x))) +
        geom_col(position = "dodge") +
        scale_x_continuous(limits = c(0, 6)) +
      scale_y_continuous(labels = scales::number_format()) +
        labs(y = "",
             x = "")
  }
) |> 
  reduce(`/`)
```


```{r}
#| fig-width: 6
#| fig-height: 8

map(
  c("donor_class", "donation_type", "gender"),
  \(x) {
    data |> 
      filter(number_of_donations < 6) |> 
      summarise(
        n = n(),
        .by = c(!!sym(x), number_of_donations)
      ) |>
      # mutate(n = n / n(), .by = !!sym(x)) |> 
      ggplot(aes(x = number_of_donations, y = n, fill = !!sym(x))) +
      geom_col() +
      facet_wrap(vars(!!sym(x))) +
      scale_y_continuous(labels = scales::number_format()) +
      labs(y = "",
           x = "",
           subtitle = x
           ) +
      theme(legend.position = "none")
  }
) |> 
  reduce(`/`)
```

## How many times has a donor donated in last years?

Add a column with the cumsum of the number of donations and expand the data frame adding the no donations for whom donated and for next years no more.

```{r}
data |> 
  arrange(unique_number, age) |> 
  mutate(
    total_donations = cumsum(number_of_donations),
    .by = unique_number,
    .after = number_of_donations
  ) -> data

```



```{r}
data |>
  slice_max(total_donations, by = unique_number) |> 
  filter(age < 70, age > 15) |> 
  ggplot(aes(x = age, y = total_donations)) +
  # geom_point(position = "jitter") +
  geom_hex(aes(fill = class_year), alpha = .4) +
  geom_smooth(aes(color = class_year), linetype = "dashed", method = "lm") +
  guides(color = "none")
```

```{r}
data |> 
  ggplot(aes(x = total_donations, y = after_stat(density))) +
  geom_histogram(bins = 50) +
  scale_x_sqrt(breaks = c(0, 1, 2, 5, 10, 25, 50, 100, 200)) +
  labs(
    title = "Distribuition of total donations",
    subtitle = "The x axis is scaled with a square root transformation"
  )
```


Maybe I should slice by max donations by person. Other option is just to slice max by year that is the last year of observation that we have. But it's not the same year for each one.

```{r}
data |> 
  slice_max(total_donations, by = unique_number) -> data_model

data_model |> 
  ggplot(aes(x = total_donations, y = ..density..)) +
  geom_histogram(bins = 50) +
  scale_x_sqrt(breaks = c(0, 1, 2, 5, 10, 25, 50, 100, 200)) +
  labs(
    title = "Distribuition of total donations in the last year of donations of each donator",
    subtitle = "The x axis is scaled with a square root transformation"
  )

```



### Poisson

```{r}
mod <- glm(total_donations ~ age + class_age + first_donation_year + gender,
          data = data_model |> filter(total_donations < 100),
          family = poisson(link = "log"),
          # family = Gamma(link = "inverse"),
          )
summary(mod)
plot(mod)
```
### Poisson

```{r}
mod <- glm(total_donations ~ age + class_age + first_donation_year + gender,
          data = data_model |> filter(total_donations < 100),
          family = quasipoisson(link = "log"),
          # family = Gamma(link = "inverse"),
          )
summary(mod)
plot(mod)
```


### Tweedie

```{r, eval=FALSE}
library(statmod)
library(tweedie)
library(doParallel)

data_filtered <- data_model |> filter(total_donations < 100)

data_filtered$gender <- factor(data_filtered$gender)
data_filtered$class_age <- factor(data_filtered$class_age)

# Set up parallel backend
numCores <- detectCores() - 1  # Use one less than the total cores
cl <- makeCluster(numCores)
registerDoParallel(cl)

# Function to estimate power parameter in parallel
tweedie_tuning <- foreach(p = seq(.5, 3, length.out = numCores), .combine = c, .packages = 'tweedie') %dopar% {
  tweedie.profile(
    total_donations ~ class_age + gender,
    data = data_filtered |> dplyr::slice_sample(n = 1e3, by = c(class_age, gender)),
    p.vec = p,
    do.plot = FALSE
  )
}

# Stop the cluster after usage
stopCluster(cl)
registerDoSEQ()

# Assuming `tweedie_tuning` is your original list
tweedie_tuning_unlisted <- tweedie_tuning |> 
  unlist(use.names = TRUE)

# Convert to a data frame
tweedie_tuning_df <- tibble(
  name = names(tweedie_tuning_unlisted),
  value = tweedie_tuning_unlisted
)

# Group by name and create a wide format data frame
tweedie_tuning_wide <- tweedie_tuning_df |> 
  group_by(name) |> 
  mutate(row = row_number()) |> 
  pivot_wider(names_from = name, values_from = value) |> 
  select(-row) 

tweedie_tuning_wide |>
  mutate(across(c(p, phi), as.numeric)) |> 
  slice_max(phi) |> 
  pull(p) -> best_p

best_p
```
```{r}
best_p <- 1.19
```


```{r}
mod_tweedie <- glm(total_donations ~ age + class_age + first_donation_year + gender,
                   data = data_filtered,
                   family = tweedie(var.power = best_p, link.power = 0))

summary(mod_tweedie)

plot(mod_tweedie)
```





### Gamma

```{r}
mod_gamma <- glm(total_donations ~ age + class_age + first_donation_year + gender,
          data = data_model |> filter(total_donations < 100),
          family = Gamma(link = "inverse"),
          )
summary(mod_gamma)
plot(mod_gamma)
```


```{r}
library(tidymodels)
mod_gamma |> 
  tidy() |> 
  ggplot(aes(y = term, x = estimate)) + 
  geom_point(color = "springgreen") +
  geom_vline(xintercept = 0, color = "tomato") +
  geom_errorbar(aes(xmin = estimate - 1.96 * std.error, xmax = estimate + 1.96 * std.error),
                width = .2)
```
```{r}
ggstatsplot::ggcoefstats(mod_gamma)
```


## Model on the last year

The Idea is to make a predictable model to provide a prediction for the next year based on the past blood donation of a donor.
In this case, we're not taking into account the donors who in the future won't donate and the donors who in the past didn't donate.

The dataframe will be structured in one row for each donor with the information about the past donations.

### Pivot the data

We have to decide how to manage the different donation type. We could analyze jointly or separately.
Now, we restrict the analysis to blood donation (`SANGUE`) and donor class equal to *P*.

```{r}
data |> 
  filter(donation_type == 'SANGUE', donor_class == 'P') |> 
  pivot_wider(
    names_from = year,
    names_prefix = "y_",
    values_from = number_of_donations,
    id_cols = unique_number,
    values_fill = 0
  ) |> 
  # take who has donated in the last two year
  filter(if_any(c(y_2022, y_2021), \(x) x > 0)) -> donations
donations
```

Join with the sociodemographic data

```{r}
data |> 
  reframe(
    class_year,
    birth_year,
    first_donation_year,
    gender,
    .by = unique_number
  ) |> 
  distinct() -> sociodemographic

right_join(
  sociodemographic,
  donations,
  by = "unique_number"
) -> recent_donations

recent_donations
```

Plot the data

```{r}
recent_donations |> 
  summarise(
    n = n(),
    .by = c(class_year, y_2022)
  ) |> 
  ggplot(aes(x = y_2022, y = class_year, fill = n)) +
  geom_tile()
```

```{r}
recent_donations |> 
  summarise(
    n = n(),
    .by = c(birth_year, y_2022)
  ) |> 
  ggplot(aes(x = y_2022, y = birth_year, z = n)) +
  geom_contour_filled()
```
```{r}
recent_donations |> 
  ggplot(aes(x = y_2022)) +
  geom_histogram()
```
During the covid the correlation has significatly decreased 

```{r}
recent_donations |> 
  select(paste0("y_", 2010:2023)) |> 
  cor() |> 
  as_tibble() |> 
  bind_cols(var_1 = paste0("y_", 2010:2023)) |> 
  pivot_longer(-var_1, names_to = "var_2", values_to = "correlation") |> 
  ggplot(aes(x = var_1, y = var_2, fill = correlation)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1, 1), space = "Lab", 
                       name="Pearson\nCorrelation") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, 
                                   size = 12, hjust = 1)) +
  coord_fixed()
```

```{r}
glm(y_2019 ~ y_2018 + y_2017 + y_2016 + gender + birth_year * first_donation_year, data = recent_donations,
    family = quasipoisson()
    ) -> model_pre_covid 

model_pre_covid |> 
  summary()
```
In the Q-Q Residuals plot, could the strange curve around 0 be caused from the underdispersion parameter of the quasipoisson?

```{r}
plot(model_pre_covid)
```
```{r}
gtsummary::tbl_regression(
  model_pre_covid,
  exponentiate = T
)
```
```{r}
donations |> 
  select(-unique_number) |> 
  pivot_longer(everything()) |> 
  ggplot(aes(x = value)) +
  geom_histogram(bins = 50) 
```


```{r}
donations |> 
  select(-unique_number) |> 
  pivot_longer(everything()) |> 
  tabyl(value) |> 
  adorn_percentages(denominator = "col") |> 
  adorn_pct_formatting() 
```

