---
title: "Explorative Analysis"
format: 
  html:
    toc: true
    code-fold: true
    df-print: paged
    warnings: false
---

## Load Libraries

```{r}
pacman::p_load(
  tidyverse,  # A set of many useful libraries
  readxl,     # To import the dataset from Excel
  here,       # To avoid problems with file directories
  janitor,    # To clean data in a fast way
  gt,         # Output tables
  gtsummary,  # Output tables for models and survival data
)
```


Set ggplot theme

```{r}
theme_set(theme_minimal())
```


## Import Data

```{r}
data <- read_csv(
  here("data", "FINAL", "dataframe_cleaned.csv")
  )

data |> 
  slice_sample(n = 1, by = c(donation_type, gender)) |> 
  gt() |> 
  tab_header("Sample of the dataset cleaned",
             "The sample was stratified by the variables donation_type and gender")
```

## Some Checks

Check the unique donors number

```{r}
data |> 
  distinct(unique_number) |> 
  nrow()
```
Check if there are any duplicate rows

```{r}
data |> 
  janitor::get_dupes()
```
### Clean the df

```{r}
data |> 
  distinct() |> 
  mutate(
    class_year = cut(birth_year, 
                     breaks = seq(1900, 2010, by = 10), 
                     dig.lab = 4,
                     include.lowest = TRUE
                     ),
    class_age = cut(age, 
                     breaks = c(seq(0, 70, by = 10), max(age)), 
                     dig.lab = 3,
                     include.lowest = TRUE
                     ),
    .before = birth_year
  ) -> data
```



## Some analysis

Age of donors for the last year

```{r}
data |> 
  filter(
    year == max(year) | year == min(year)
  ) |> 
  ggplot(aes(x = gender, y = age, fill = gender)) +
  geom_violin() +
  facet_wrap(~year) +
  labs(
    title = "Donor age distribution in 2009 and 2023 by gender"
    )
```


```{r}
data |> 
  select(class_year, donation_type) |> 
  tbl_summary(
    by = donation_type
  ) |> 
  add_overall()
```

```{r}
data |> 
  reframe(
    class_year, donor_class,
    .by = unique_number
  ) |> 
  select(class_year, donor_class) |> 
  tbl_summary(
    by = donor_class
  ) |> 
  add_overall()
```
Maybe the donations are reffered to the year. But is it possible to donate 17 times per year?

```{r}
data |> 
  tabyl(number_of_donations, gender) |>
  adorn_percentages(denominator = "col") |>  
  adorn_pct_formatting()
```

## Time series

```{r}
data |>
  filter(
    donation_type == "SANGUE",
    birth_year > 1930
    ) |> 
  summarise(
    donations = sum(number_of_donations),
    .by = c(class_year, year)
  ) |> 
  ggplot(aes(year, donations, color = class_year, group = class_year)) +
  geom_line() +
  scale_color_brewer() +
  labs(
    title = "Number of donations by year and class of birth years"
  )
```
```{r}
data |>
  filter(
    donation_type == "SANGUE",
    birth_year > 1930
    ) |> 
  summarise(
    donations = sum(number_of_donations),
    .by = c(class_age, year)
  ) |> 
  ggplot(aes(year, donations, color = class_age, group = class_age)) +
  geom_line() +
  scale_color_brewer(direction = -1) +
  labs(
    title = "Number of donations by year and class of birth years"
  )
```



## Integrate Data


Integrate data with Trieste residents to get more insights

```{r}
residenti <- read_csv(here("data", "residenti_Trieste_e_Gorizia.csv"))

residenti |> 
  filter(
    Sesso != "totale"
  ) |> 
  rename(
    age = Età,
    gender = Sesso,
    year = TIME,
    population = Value,
  ) |> 
  summarise(
    across(population, sum),
    .by = c(age, gender, year)  
  ) |> 
  mutate(
    age = str_extract(age, "[:digit:]*") |> as.numeric(),
    gender = case_when(
      gender == "maschi" ~ "M",
      gender == "femmine" ~ "F",
      T ~ NA
    )
  ) -> residenti
```


### Estimating Past Residents

For the years prior to 2019, one could use the 2019 data from ISTAT, adjust the ages, and account for population changes with mortality tables, assuming a constant flow of immigrants and emigrants.

```{r}
lifecontingencies::demoIta |> 
  transmute(
    age = X,
    maschi = SIM02,
    femmine = SIF02,
    across(c(maschi, femmine), \(x) x / lag(x, default = 1e5), .names = "{col}_px")
  ) |> 
  pivot_longer(ends_with("px"), names_to = "gender", values_to = "px") |>
  # select(-maschi, - femmine) |> 
  mutate(
    gender = case_when(
      gender == "maschi_px" ~ "M",
      gender == "femmine_px" ~ "F",
      T ~ NA
    )
  ) -> life_table

residenti |> 
  add_row(year = 2009:2018, .before = 1) |> 
  complete(age, gender, year) |> 
  filter(!if_any(c(age, gender), is.na)) |> 
  left_join(life_table, by = c("age", "gender")) |> 
  arrange(-year, age) |>
  filter(gender == "M") |> 
  pivot_wider(names_from = year, values_from = population, names_prefix = "y_") |> 
  mutate(
    y_2018 = lead(y_2019, default = 0) * maschi / lead(maschi, default = 0),
    y_2017 = lead(y_2019, n = 2, default = 0) * maschi / lead(maschi, default = 0),
    y_2016 = lead(y_2019, n = 3, default = 0) * maschi / lead(maschi, default = 0),
  )
```

Let $l_x$ be the population from the mortality table at age $x$, and let $n_x^y$ be the resident population in Trieste in year $y$ at age $x$. To estimate the population in year $y_j$ at age $x$, given the population in year $y_i$, the following formula is used:

$n^{y_i}_x = n^{y_j}_{x - (y_i - y_j)} \cdot \frac{l_x}{l_{x - (y_i - y_j)}}$

```{r}
filled_residenti <-
  residenti |> 
  add_row(year = 2009:2018, .before = 1) |> 
  complete(age, gender, year) |> 
  filter(!if_any(c(age, gender), is.na)) |> 
  left_join(life_table, by = c("age", "gender")) |> 
  arrange(-year, age) |>
  pivot_wider(names_from = gender, values_from = c(px, population)) |> 
  pivot_wider(names_from = year, values_from = c(population_M, population_F))

years <- 2018:2009
gender <- "M"

filled_residenti <- reduce(years, function(df, year) {
  col <- paste0("population_", c("F", "M"), "_", year)
  col_19 <- paste0("population_", c("F", "M"), "_2019")
  df |> 
    mutate(
      !!sym(col[1]) := lead(!!sym(col_19[1]), 2019 - year, 0) * femmine / lead(femmine, 2019 - year, 0),
      !!sym(col[2]) := lead(!!sym(col_19[2]), 2019 - year, 0) * maschi / lead(maschi, 2019 - year, 0)
      )
}, .init = filled_residenti) |> 
  mutate(across(starts_with("population"), round))

filled_residenti
```

```{r}
filled_residenti |> 
  select(age, starts_with("population")) |> 
  pivot_longer(starts_with("population"), values_to = "population") |> 
  separate(name, sep = "_", into = c("pop", "gender", "year")) |> 
  select(-pop) |> 
  mutate(across(year, as.numeric)) -> residenti
```


### Plot Integrated Data

```{r}
data |> 
  left_join(residenti, by = c("gender", "year", "age")) |>
  filter(year %% 4 == 3) |>
  summarise(
    ratio_donors = n() / first(population),
    .by = c(gender, year, age)
  ) |> 
  filter(age < 75) |> 
  ggplot(aes(x = age, y = ratio_donors, fill = gender, alpha = year)) +
  geom_col() +
  facet_grid(rows = vars(year), cols = vars(gender)) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_alpha_continuous(range = c(.4, 1)) +
  labs(
    title = "Donors population ratio by gender and year"
  ) +
  theme(legend.position = "none")
```


```{r}
data |> 
  left_join(residenti, by = c("gender", "year", "age")) |>
  reframe(
    n = n(),
    .by = c(gender, year, age, class_age, population)
  ) |> 
  summarise(
    ratio_donors = sum(n) / sum(population),
    .by = c(year, gender, class_age)
  ) |> 
  ggplot(aes(x = year, y = ratio_donors, color = class_age)) +
  geom_line() +
  facet_grid(cols = vars(gender)) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Donors population ratio by gender and year"
  )
```



